### 2024-03-04 docs/manuscript/manuscript.qmd

HRV was recorded through a non-invasive chest band, immediately before, during, and immediately after the execution of the two-minutes step test (TMST). Vital signs, including blood pressure, were monitored throughout the test, and participants' well-being was visually checked to ensure they were comfortable and mentally prepared. For resting HRV measurements (prior and posterior to TMST), the volunteers remained seated in a chair, while proper support of the feet and back, and avoiding talking during the recording was ensured. RR intervals were recorded continuously during the last 10 min of rest, and were subsequently analyzed for 5 min in each occasion. The breathing rate was spontaneous. As part of the protocol, it was ensured that the participant had a blood pressure less than 140/90 mmHg to start the HRV measurements.

@@

Physical and Physiological assessments, as well as the initial interview, were made by professional physiotherapists.

@@

Cardiac autonomic activity was assessed using RR interval recordings (HRV) obtained through the Polar Team2 system (Polar®) application. Artifacts and ectopic heartbeats, which did not exceed 3% of the recorded data, were excluded [@malik1996heart]. Time-domain parameters considered for analysis included the square root of the mean squared differences of successive RR intervals (RMSSD, expressed in ms) as an index of parasympathetic activity [@pmid:20033207], and the standard deviation of RR intervals (SDNN), reflecting total variability encompassing both sympathetic and parasympathetic contributions to cardiac autonomic function [@pmid:9401419; @buchheit2006cardiac]. Additionally, the Stress Index (SI) and Parasympathetic and Sympathetic Nervous System Index (PNS and SNS, respectively) were calculated. The PNS Index, indicative of total vagal stimulation, was derived from mean R-R intervals, RMSSD, and Poincaré Plot Index SD1 in normalized units (linked to RMSSD), representing deviations from normal population averages [@pmid:9401419; @GSID:XkAQwTGLgNsJ]. The SNS Index, reflecting total sympathetic stimulation, was derived from mean R-R intervals, Baevsky’s Stress Index (positively related to cardiovascular system stress and cardiac sympathetic activity), and the Poincaré Plot Index SD2 in normalized units (related to SDNN) with interpretation similar to the PNS Index [@pmid:9401419; @GSID:XkAQwTGLgNsJ]. The SI serves as an indicator representing the ANS control system's workload [@doi:10.1159/000513781], normalized by the square root of Baevsky’s SI [@2009MethodicalRU]. All analyses were conducted using Kubios HRV® software (Kuopio, Finland).

<!--- ESTA VARIABLE NO ES MENCIONADA EN LOS RESULTADOS, ES CUESTIONABLE SU MENCIÓN EN EL PAPER --->
### Two-Minutes Step Test **RECONSIDERAR ESTA VARIABLE**

The TMST is a sub-test from the Senior Fitness Test, demanding a low to moderate intensity [@Rikli1999; @ijerph20054456]. The test consists of a two-minute assessment designed to evaluate cardiorespiratory fitness. Participants are instructed to raise their right knee to a marked point on a wall as many times as possible within the given time frame, ensuring that each raise reaches at least a 70° angle at the thigh-femoral joint.
<!--- ESTA VARIABLE NO ES MENCIONADA EN LOS RESULTADOS, ES CUESTIONABLE SU MENCIÓN EN EL PAPER --->

@@

As a way of controlling for the influence of psycho-physiological variables on the cardiac autonomic response of the subjects to exercise, we assess the average causal mediation effect (ACME), the average direct effect (ADE) of the main effects after taking into account the effect of moderator variables into the observed relationships and the proportion of the effect that its mediated by these variables [@kosuke2010identification]. To assess the significance of moderating variables, we used non parametric bootstrapping by using Monte Carlo resampling, and bias-corrected and accelerated CI~95%~, using the *mediation* R package to this end [@tingley2014mediation].

We defined a type I error rate at 5% ($\alpha$ = 0.05) as our threshold for null hypothesis significance testing, and non-overlapping CI~95%~ over the null effect for sex differences. All analyses were computed using the *R* programming language for statistical computing, [@rlanguage].

@@

<!---
#> cc_grasa_total_porcentaje, hrv_rmssd_peri, r = 0.253, p = 0.029
#> cc_grasa_total_porcentaje, hrv_rmssd_peri, rho = 0.249, p = 0.032
#> 
#> cc_grasa_total_porcentaje, hrv_stress_peri, r = -0.258, p = 0.027
#> cc_grasa_total_porcentaje, hrv_stress_peri, rho = -0.282, p = 0.015
#> 
#> cc_grasa_total_porcentaje, hrv_sns_peri, r = -0.252, p = 0.03
#> cc_grasa_total_porcentaje, hrv_sns_peri, rho = -0.257, p = 0.027
#> 
#> cc_grasa_total_porcentaje, hrv_sdnn_peri, r = 0.269, p = 0.02
#> cc_grasa_total_porcentaje, hrv_sdnn_peri, rho = 0.284, p = 0.014
--->

![](figures/fig-1.jpeg)
**Figure 1**. Main correlations among psychological, cardiac autonomic and performance variables. Non-significant correlations (i.e., p > 0.05) were crossed with and "X". The more intense the color the stronger the correlation, either positive or negative.

<!--- Figura con cuatro mostrando las correlaciones --->

@@

In addition, a greater body fat percentage was negatively correlated with the SPPB total score (r = -0.273, p = 0.015), menawhile, no other metrics of body composition were associated with physical performance of the SPPB (i.e., p > 0.05).

<!---
#> cc_grasa_total_porcentaje, sppb_total, r = -0.273, p = 0.015
#> cc_grasa_total_porcentaje, sppb_total, rho = -0.25, p = 0.026
--->

@@

Physical fitness, reflected through SPPB total score, was positively correlated with a greater sympathetic activity during exercise, which was reflected by SNS index (r = 0.313, p = 0.006), mean HR (r = 0.355, p = 0.002) and its inverse counterpart mean R-R (r = -0.35, p = 0.002).

<!--- 
#> hrv_sns_peri, sppb_total, r = 0.313, p = 0.006
#> hrv_sns_peri, sppb_total, rho = 0.332, p = 0.004

#> hrv_mean_hr_peri, sppb_total, r = 0.355, p = 0.002
#> hrv_mean_hr_peri, sppb_total, rho = 0.375, p = 0.001

#> hrv_mean_rr_peri, sppb_total, r = -0.35, p = 0.002
#> hrv_mean_rr_peri, sppb_total, rho = -0.38, p = 0.001

--->

However, this appears to be accompanied by an proportional decrease in parasympathetic drive during exercise with increasing levels of physical fitness, which was denoted by the inverse correlation of the later with PNS index during (r = -0.375, p = 0.001) and after exercise (r = -0.229, p = 0.049), in addition to RMSSD during exercise (r = -0.294, p = 0.010)

<!--- 
#> hrv_pns_peri, sppb_total, r = -0.375, p = 0.001
#> hrv_pns_peri, sppb_total, rho = -0.372, p = 0.001

#> hrv_pns_post, sppb_total, r = -0.229, p = 0.049
#> hrv_pns_post, sppb_total, rho = -0.238, p = 0.04

#> hrv_rmssd_peri, sppb_total, r = -0.294, p = 0.01
#> hrv_rmssd_peri, sppb_total, rho = -0.282, p = 0.014
--->

<!--- Figura 2 --->

## Mediation

[Pendiente]

@@

This study aimed to investigate the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older people. Through the correlational analysis, we aimed to elucidate the  interplay between these factors and their implications for cardiac autonomic function during physical efforts.

It was hypothesized that a higher body fat percentage would be inversely related to  physical aptitude. Our results aligns with this hypothesis and previous research regarding body composition changes and physical performance decrease during aging [@pmid:35832415; @pmid:36256894].

Furthermore, the increase in body fat percentage was related to a reduced sympathetic activity response and a higher parasympathetic activity during TMST, as hypothesized (i. e., SNS index, SI, SDNN and RMSSD). Throughout the course of physical efforts, ANS is expected to shift into a sympathetic predominance, ensuring a sufficient response to higher metabolic demands [@pmid:33246163]. However, aged individuals may present an impaired ANS response to physical exercise [@ijerph20054456]. In that sense, our results support that body composition may play a role in that response, opening opportunities for targeted interventions.

Nonetheless, the moderating effect of physical fitness on the relationship proposed above, suggests that more physically active individuals may counteract the negative effect of a high body fat percentage on cardiac autonomic response to exercise. This result highlights the importance of promoting a physically active lifestyle and favor the maintenance of physical capacity, regardless of nutritional status [@pmid:33402165; @pmid:30002462].

Regarding physical fitness and cardiac autonomic response to exercise, as expected, our results evidenced that individuals with a higher sympathetic activity during exercise, as measured by SNS index, had a greater physical fitness while the opposite observation was made for parasympathetic activity (i. e., PNS index and RMSSD). Previous research had demonstrated the utility of HRV decrease as an index of cardiorespiratory fitness [@pmid:36054204], supporting the hypothesis that a higher sympathetic activity facilitates the performance during physical moderate effort. Even more, it has been previously demonstrated that physically active individuals may have an enhanced autonomic response to exercise, strengthening this relationship [@pmid:28721030; @pmid:24799513].

Our results demonstrate the interplay between physical fitness, body fat, and cardiac autonomic response to exercise in older individuals. Physical fitness is negatively affected by higher body fat percentage , while also being associated with impaired cardiac autonomic response to exercise, characterized by decreased sympathetic activity and increased parasympathetic tone. Additionally, this altered autonomic response is related to lower physical performance in older adults, both variables being important predictors of health outcomes in this population [@pmid:30627571; @pmid:30104034]. Overall, our findings underscore the triangular relationship among these variables.

However, our results did not evidence a moderating effect of GDS-30 and BAI scores, challenging initial hypothesis. This finding suggests a nuanced interplay between psychological variables and physical/physiological measures in older people, underscoring the need for further research to elucidate underlying mechanisms and variables beyond depression and anxiety. While the lack of moderating effects does not diminish the significance of psychological factors in aging, it highlights the complexity of their influence.

In addition to the potential strengths of this cross-sectional study, it is not without limitations. In first place, our study design prevents us from making objective inferences about the direction of the relationships, highlighting the relevance of longitudinal research exploring causal relations between our variables and underlying mechanisms. In second place, a relatively small sample was included, limiting generalizability and statistical power of the analysis. Furthermore, the sample consisted mainly of women (82.7%), limiting our ability to make sex comparisons and decreasing the representativeness of the sample. It should also be noted that the relationships between physical fitness and cardiac autonomic response to exercise were not made within the same instrument, due to the possible lack of SPPB to generate an adequate physical stress, and of TMST to inform general physical fitness of the participants. Further research could investigate the impact of cardiac autonomic response on performance during the same task, in this age group.

@@

The results highlighted a triangular relationship between the variables. Body fat percentage was importantly related to both physical fitness and cardiac autonomic activity during exercise, while sympathetic activity was associated with a grater physical fitness in older people. Furthermore, physical fitness may moderate the associations between body fat and cardiac autonomic activity, highlighting its importance in older people's health. However, further research is needed to elucidate moderating effects of psychological variables in this field.

>>>>

HRV was recorded through a non-invasive chest band, immediately before, during, and immediately after the execution of the two-minutes step test (TMST). Vital signs, including blood pressure, were monitored throughout the test, and participants' well-being was visually checked to ensure they were comfortable and prepared. For resting HRV measurements (before and after the application of the TMST), the volunteers remained seated in a chair, with feet and back supported, ensuring avoiding to talk during the recordings. R-R intervals were recorded continuously during the last 10 min of rest, and were subsequently analyzed for 5 min in each occasion. The breathing rate was spontaneous. As part of the protocol, it was ensured that the participant needed to have a blood pressure less than 140/90 mmHg to start the HRV measurements.

@@

Physical and physiological assessments, as well as the initial interview, were made by professional physiotherapists.

@@

Cardiac autonomic activity was assessed using R-R interval recordings (HRV) obtained through the Polar Team2 system (Polar®) application. Artifacts and ectopic heartbeats, which did not exceed 3% of the recorded data, were excluded [@malik1996heart]. Time-domain parameters considered for analysis included the square root of the mean squared differences of successive R-R intervals (RMSSD, expressed in ms) as an index of parasympathetic activity [@pmid:20033207], and the standard deviation of RR intervals (SDNN), reflecting total variability encompassing both sympathetic and parasympathetic contributions to cardiac autonomic function [@pmid:9401419; @buchheit2006cardiac]. Additionally, the Stress Index (SI) and Parasympathetic and Sympathetic Nervous System Index (PNS and SNS, respectively) were computed. The PNS Index, indicative of total vagal stimulation, was derived from mean R-R intervals, RMSSD, and Poincaré Plot Index SD1 in normalized units (linked to RMSSD), representing deviations from normal population averages [@pmid:9401419; @GSID:XkAQwTGLgNsJ]. The SNS Index, reflecting total sympathetic stimulation, was derived from mean R-R intervals, Baevsky’s Stress Index (positively related to cardiovascular system stress and cardiac sympathetic activity), and the Poincaré Plot Index SD2 in normalized units (related to SDNN) with interpretation similar to the PNS Index [@pmid:9401419; @GSID:XkAQwTGLgNsJ]. The SI serves as an indicator representing the ANS control system's workload [@doi:10.1159/000513781], normalized by the square root of Baevsky’s SI [@2009MethodicalRU]. All analyses conducted to compute HRV related indices were performed using Kubios HRV® software (Kuopio, Finland).

### Two-Minutes Step Test

The TMST is a sub-test from the Senior Fitness Test, demanding a low to moderate intensity [@Rikli1999; @ijerph20054456]. The test consists of a two-minute assessment designed to evaluate cardiorespiratory fitness. Participants are instructed to raise their right knee to a marked point on a wall as many times as possible within the given time frame, ensuring that each raise reaches at least a 70° angle at the thigh-femoral joint. The number of valid steps were recorded for each subject.

@@

As a way of controlling for the influence of psycho-physiological variables on the cardiac autonomic response of the subjects to exercise, we assess the average causal mediation effect (ACME), the average direct effect (ADE) of the main effects after taking into account the effect of moderator variables into the observed relationships and the proportion of the effect that its mediated by these variables [@kosuke2010identification]. To assess the significance of moderating variables, we used non overlapping CI~95%~, estimated based on non parametric bootstrapping by using Monte Carlo resampling, and bias-corrected and accelerated CI~95%~, using the *mediation* R package to this end [@tingley2014mediation].

We defined a type I error rate at 5% ($\alpha$ = 0.05) as our threshold for null hypothesis significance testing, and non-overlapping CI~95%~ over the null effect for sex differences and mediation analyses. All analyses were computed using the *R* programming language for statistical computing, [@rlanguage].

@@

![](figures/fig-1.jpeg)
**Figure 1**. Main correlations among psychological, cardiac autonomic and performance variables. Non-significant correlations (i.e., p > 0.05) were crossed with and "X". The more intense the color the stronger the correlation, either positive or negative. GDS, Geriatric Depression Scale; BAI, Beck Anxiety Inventory; SPPB, Short Physical Performance Battery; TMST, Two-Minutes Step Test; BMI, Body Mass Index.

@@

In addition, greater SPPB score was observed in leaner (r = -0.273, p = 0.015) and individuals with lower BMI values (r = -0.225, p = 0.047). Similar findings were observed between total steps from the TMST and body fat percentage (r = -0.279, p = 0.013) and bodyweight (r = -0.232, p = 0.041). Not enough evidence was observed to suggest a correlation between total muscle mass with either total SPPB score (r = -0.063, p = 0.584) or total number of steps in TMST (r = -0.070, p = 0.541).

@@

Physical fitness, reflected through SPPB total score and total steps in TMST, was positively correlated with a greater sympathetic activity during exercise, which was reflected by SNS index (SPPB score, r = 0.313, p = 0.006; TMST steps, r = 0.265, p = 0.022), mean HR (SPPB score, r = 0.355, p = 0.002; TMST steps, r = 0.338, p = 0.003) and its inverse counterpart mean R-R (SPPB score, r = -0.35, p = 0.002; TMST steps, r = -0.311, p = 0.007). Total number of steps from TMST was also inversely correlated with parasympathetic indices during exercise like RMSSD (r = -0.285, p = 0.014) and PNS index (r = -0.344, p = 0.003).

However, this appears to be accompanied by an proportional decrease in parasympathetic drive during exercise with increasing levels of physical fitness displayed in the SPPB, which was denoted by the inverse correlation of the later with PNS index during (r = -0.375, p = 0.001) and after exercise (r = -0.229, p = 0.049), in addition to RMSSD during exercise (r = -0.294, p = 0.010)

## Mediation analysis

Mediation analyses suggest many potential influential effects for the relationships for body composition and parasympathetic indices. In this context, the average direct effect (ADE) of body fat percentage on RMSSD during exercise (ADE = 0.088, CI~95%~[-0.005, 0.200], p = 0.077) was accentuated when considering the influence of SPPB score (ACME = 0.0289, CI~95%~[0.003, 0.110], p = 0.057; Total effect = 0.117, CI~95%~[0.028, 0.240], p = 0.012). Similar findings were observed when considering the mediation effect of TMST steps into account (ACME = 0.0283, CI~95%~[0, 0.07], p = 0.079; Total effect = 0.117, CI~95%~[0.028, 0.240], p = 0.012). No other parasympathetic indicators were influenced or mediated by fitness or psychological related measures.

Additionally, and in the case on sympathetic indicators, the effect of body fat on SNS index during exercise (ADE = -0.050, CI~95%~[-0.107, 0.01], p = 0.077) was influenced by SPPB score in similar way as with RMSSD, potentiating the original main observed effect (ACME = -0.016, CI~95%~[-0.049, 0], p = 0.041; Total effect = -0.066, CI~95%~[-0.125, -0.010], p = 0.016). Similar effects were observed with TMST steps in this regard (ACME = -0.015, CI~95%~[-0.042, 0], p = 0.050; Total effect = -0.066, CI~95%~[-0.125, -0.010], p = 0.016). No other sympathetic indices were notoriously modified in the presence of either SPPB score, TMST steps or psychological variables.

When assessing potential mediators for the effect of body composition on fitness related measures, we found that the effect of body fat percentage on SPPB score (ADE = -0.075, CI~95%~[-0.123, -0.020], p = 0.007) is partially diminished when considering the influence of total muscle mass (ACME = 0.011, CI~95%~[0, 0.040], p = 0.107; Total effect = -0.064, CI~95%~[-0.111, -0.010], p = 0.014). 
Similar mediation effects of total muscle mass were observed (ACME = 0.118, CI~95%~[0.005, 0.350], p = 0.075) when assessing the effect of body fat percentage on TMST steps (ADE = -0.773, CI~95%~[-1.193, -0.300], p = 0.001; Total effect = -0.655, CI~95%~[-1.084, -0.210], p = 0.004). No other mediating effects were observed for any of the psychological measures for any of the relationships between body composition and physical fitness measures.

@@

This study aimed to investigate the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older people. Through correlational analyses, we aimed to elucidate the  interplay between these factors and their implications for cardiac autonomic function during physical efforts.

It was hypothesized that a higher body fat percentage would be inversely correlated to physical fitness. Our results aligns with this hypothesis and with those of previous research regarding body composition and physical performance, with a proportional decrease in aging individuals [@pmid:35832415; @pmid:36256894].

Furthermore, greater body fat percentage was related to a reduced sympathetic drive and higher parasympathetic activity during the TMST, as hypothesized. Throughout the course of physical efforts, the ANS is expected to shift into a sympathetic state, ensuring an efficient response to greater metabolic demands [@pmid:33246163]. However, aged individuals may present an impaired ANS response to physical exercise [@ijerph20054456]. In that sense, our results support that body composition may play a pivotal role in that response, generating novel insights for potential interventions.

Nonetheless, the moderating effect of physical fitness related measures on the relationship proposed earlier, suggests that greater physical activity could potentially counteract the negative effects of a high body fat percentage on cardiac autonomic response to exercise, possibly, at the same extent as seen in leaner individuals. This result highlights the importance of promoting a physically active lifestyle and favor the maintenance of physical capacity, regardless of nutritional status [@pmid:33402165; @pmid:30002462].

Regarding the link between physical fitness and cardiac autonomic response to exercise, our results evidenced that individuals with a high sympathetic drive while exercising, had greater physical fitness. Additionally, previous research had demonstrated the utility of HRV and exercise-induced responses in those metrics as a potential marker of cardiorespiratory fitness in these individuals [@pmid:36054204], supporting the current hypothesis that a greater sympathetic drive facilitates performance and its related measures during moderate intensity physical activities. Even more, it has been previously demonstrated that physically active individuals have an enhanced autonomic response to exercise, playing a pivotal role strengthening this hypothesis [@pmid:28721030; @pmid:24799513].

Our results provide evidence to support the interplay between physical fitness, body fat, and cardiac autonomic response to exercise in older individuals. Physical fitness is negatively affected by higher body fat percentage, while also being associated with impaired cardiac autonomic response to exercise, characterized by decreased sympathetic drive and increased parasympathetic tone. Additionally, this altered autonomic response is related to lower physical performance in older individuals, both variables being important predictors of health outcomes in this population [@pmid:30627571; @pmid:30104034]. Overall, our findings underscore the triangular and bidirectional relationship among these variables.

However, our results did not evidence a moderating effect of either GDS-30 or BAI scores, challenging our initial hypothesis. These finding suggests a nuanced interplay between psychological variables and physical/physiological measures in older people, underscoring the need for further research to elucidate underlying mechanisms and variables beyond depression and anxiety. While the lack of evidence suggesting any moderating effects does not diminish the significance of psychological factors in aging, it highlights the complexity of their influence.

In addition to the potential strengths of this cross-sectional study, it is not without limitations. First, our study design prevents us from making causal inferences about the direction of the relationships, highlighting the relevance of experimental and longitudinal research exploring causal relations between our variables and potential underlying mechanisms. Second, a relatively small sample was included, limiting generalizability and statistical power of the analysis. Furthermore, the sample consisted mainly of women (82.7%), limiting our ability to make sex comparisons and decreasing the representativeness of the sample. It should also be noted that the relationships between physical fitness and cardiac autonomic response to exercise were not made within the same instrument, due to the possible lack of SPPB to generate an adequate physical stress, and of TMST to inform general physical fitness of the participants. Further research could investigate the impact of cardiac autonomic response on performance during the same task, in this age group.

@@

The results highlighted a structural bidirectional relationship among variables. Body fat percentage was inversely correlated to both physical fitness and cardiac autonomic activity during exercise, with sympathetic drive associated with physical fitness in older people. Furthermore, physical fitness may mediate the effect of body fat on cardiac autonomic activity during exercise, highlighting its pivotal role in older people's health. However, further research is needed to elucidate moderating effects of psychological variables in this field.
